\section{Cryptography}

Cryptographic crates (and majority of all cryptographic libraries, not just for Rust)
expose high level API. Developers provide inputs to this API and get back the
desired hash, encrypted text, key, etc. In context of ZkVM, also the intermediate
states are needed to populate the witness. Main hurdle during implementation
was to reuse exposed parts of used cryptographic crates, in order to collect
internal states of these primitives.

\subsection{Poseidon2}

The Poseidon2 hash function is used to commit to the IVC state. While the
\texttt{Plonky3} library provides a implementation of Poseidon2, its standard
interface treats the permutation as a black box, returning only the final digest.
To correctly constrain the Poseidon2 execution within a CCS requires access to
the intermediate states of the permutation. This is because Poseidon2 has
a 7th degree S-box, which is dependent on previous rounds, and those on ones
before them. Thus, the computation is broken down into smaller
steps, where each step verifies a transition of a one computational segment.

A custom implementation of the Poseidon2 permutation was
constructed, structurally identical to the \texttt{Plonky3} implementation but
modified to capture and return the state vector at desired boundaries:

\begin{itemize}
	\item The state after the initial MDS matrix application.
	\item The state after each of the external initial full rounds.
	\item The state after each of the internal partial rounds.
	\item The state after each of the external terminal full rounds.
\end{itemize}

\subsection{LatticeFold}\label{subsec:latticefold}

In practice, integrating LatticeFold into the ZkVM required vendoring the
\texttt{latticefold} repository and using it as a local workspace dependency
instead of consuming it purely as an upstream dependency. The main reason
was that the ZkVM needs access not only to the final prover and verifier outputs,
but also to internal values that are useful for witness generation and for
constructing CCS constraints for the in-circuit verifier. Some helper functions
and verifier-side routines in the original library were private, which made them
unavailable for reuse from the ZkVM crate. Therefore, selected pieces of
functionality had to be made public, and in a few places thin wrapper functions
were added in order to reuse the original implementation rather than duplicating
the protocol logic.

