\chapter{Implementation}\label{chap:implementation}

This chapter presents the prototype implementation of the lattice-based RISC-V
ZkVM architecture described in the Chapter~\ref{chap:design}.
The implementation covers:

\begin{itemize}
	\item A subset of rv32imac instructions sufficient to run Nth Fibonacci and validate an Ethereum block with REVM.
	\item CCS constraints for these instructions.
	\item CCS constraints for the Poseidon2 IVC step commitment.
	\item CCS constraints for NIFS verifier.
\end{itemize}

Regarding the choice of folding scheme, this work uses the NetherMind's implementation of LatticeFold
protocol rather than its successor LatticeFold+, as the latter's library
implementation was not available at the time of system design and development.

\input{content/implementation/tech-stack.tex}

\input{content/implementation/crypto.tex}

\input{content/implementation/riscv-emul.tex}

\input{content/implementation/ccs.tex}

\input{content/implementation/optimizations.tex}

\section{High level overview}

Figure~\ref{fig:execution_flow_overview} illustrates the complete execution flow. The
setup phase prepares the VM by loading the ELF binary and constructing the CCS
constraint system. Within the execution loop, four main modules work together:
the \texttt{riscvm} module handles instruction execution and trace generation,
the \texttt{commitments} module creates Poseidon2 hashes for state components,
the \texttt{ivc} module arithmetizes the trace into the CCS witness vector,
and the \texttt{main} module performs the folding operations that
compress each step into the accumulator. The dashed line on the right shows the
recursive nature of the computation, where the \texttt{IVCStepOutput} from step
$i$ becomes the \texttt{IVCStepInput} for step $i+1$.

\begin{tikzFigure}
	\begin{tikzpicture}[
			node distance=1.8cm and 1.0cm,
			font=\sffamily\small,
			% --- Styles ---
			% Functions / Logic
			func/.style={
					rectangle,
					draw=red!80!black,
					fill=red!10,
					thick,
					minimum height=0.7cm,
					minimum width=4.5cm,
					align=center,
					rounded corners=2pt
				},
			% Data Structures
			data/.style={
					rectangle,
					draw=blue!80!black,
					fill=blue!5,
					thick,
					minimum height=0.6cm,
					minimum width=4cm,
					align=center,
					dashed,
					font=\sffamily\footnotesize
				},
			% Module Containers
			module/.style={
					draw=gray!40,
					dashed,
					inner sep=0.5cm,
					rounded corners=5pt,
					fill=white,
					fill opacity=0.0
				},
			% Module Label
			modlabel/.style={
					font=\bfseries\scriptsize,
					text=gray!60!black,
					anchor=north east
				},
			% Lines
			line/.style={
					draw,
					-Latex,
					thick,
					gray!70!black
				}
		]

		% --- 1. SETUP PHASE (Top) ---
		\node[func] (setup_vm) {\texttt{VM::load\_elf}};
		\node[func, right=0.5cm of setup_vm] (setup_ccs) {\texttt{CCSBuilder::create\_riscv\_ccs}};

		% --- 2. THE LOOP  ---

		% --- Module: riscvm ---
		\node[func, below=1.8cm of setup_vm, xshift=2.5cm] (vm_exec) {\texttt{VM::fetch\_execute}};
		\node[data, below=0.3cm of vm_exec] (trace) {\texttt{ExecutionTrace}};

		% --- Module: commitments ---
		\node[func, below=1.5cm of trace] (committer) {
			\textbf{ZkVmCommitter methods:} \\
			ivc\_step\_comm,\\
			state\_i\_comm,\\
			acc\_comm,\\
			vm\_regs\_comm,\\
			vm\_mem\_comm
		};
		\node[data, below=0.3cm of committer] (comms) {Commitments represented by\\\texttt{GoldilocksComm}\\(4 Goldilocks field elements)};

		% --- Module: ivc ---
		\node[data, below=1.5cm of comms] (ivc_input) {\textbf{\texttt{IVCStepInput}} \\ (Prev Acc + Trace + Comms)};
		\node[func, below=0.3cm of ivc_input] (arith) {\texttt{arithmetize(input, layout)}};
		\node[data, below=0.3cm of arith] (z_vec) {\texttt{Vec<GoldilocksRingNTT>}\\($z$ vector)};

		% --- Module: main ---
		\node[func, below=1.5cm of z_vec] (commit_wit) {\texttt{commit(z, ccs, scheme)}};
		\node[data, below=0.3cm of commit_wit] (cccs) {\texttt{CCCS} ($cm_i$) + \texttt{Witness} ($w_i$)};

		\node[func, below=0.3cm of cccs] (nifs) {\texttt{fold(ccs, scheme, acc, w\_acc, cm\_i, w\_i)}};
		\node[data, below=0.3cm of nifs] (output) {\textbf{\texttt{IVCStepOutput}} \\ (New Accumulator + Proof)};

		% --- 3. MODULE BOUNDARIES ---
		\begin{scope}[on background layer]
			% VM Box
			\node[module, fit=(vm_exec) (trace)] (box_riscv) {};
			\node[modlabel] at (box_riscv.north east) {\texttt{mod riscvm}};

			% Comm Box
			\node[module, fit=(committer) (comms)] (box_comm) {};
			\node[modlabel] at (box_comm.north east) {\texttt{mod commitments}};

			% IVC Box
			\node[module, fit=(ivc_input) (arith) (z_vec)] (box_ivc) {};
			\node[modlabel] at (box_ivc.north east) {\texttt{mod ivc}};

			% Latticefold Box
			\node[module, fit=(commit_wit) (cccs) (nifs) (output)] (box_fold) {};
			\node[modlabel] at (box_fold.north east) {\texttt{crate latticefold}};

			% Main Loop Container
			\node[draw=gray!80, thick, inner sep=25pt, fit=(box_riscv) (box_fold), label={[anchor=north west, font=\bfseries]north west:Execution Loop (\texttt{vm.run})}] (main_loop) {};
		\end{scope}

		% --- 4. CONNECTIONS ---

		% Init -> Loop
		% Manual orthogonal lines
		\draw[line] (setup_vm.south east) -- (vm_exec.north);
		\draw[line] (setup_ccs.south east) -- ++(0,-4.5) -- (commit_wit.east);

		% Flow down
		\draw[line] (vm_exec) -- (trace);
		\draw[line] (trace) -- (committer);
		\draw[line] (committer) -- (comms);
		\draw[line] (comms) -- (ivc_input);
		\draw[line] (ivc_input) -- (arith);
		\draw[line] (arith) -- (z_vec);
		\draw[line] (z_vec) -- (commit_wit);
		\draw[line] (commit_wit) -- (cccs);
		\draw[line] (cccs) -- (nifs);
		\draw[line] (nifs) -- (output);

		\draw[line, dashed] (output.east) -- ++(2.9, 0)
		-- ++(0, 7.3)
		-- (ivc_input.east);

		% Label for loop
		\node[font=\scriptsize\bfseries, text=gray!80!black, rotate=90] at ($(output.east) + (2.7, 3.5)$) {Accumulator ($i \to i+1$)};
	\end{tikzpicture}
	\caption{High-level overview of the implemented execution flow.}
	\label{fig:execution_flow_overview}
\end{tikzFigure}

\pagebreak

\subsection{Toolchain summary}

The implementation is built with the following software stack:

\begin{itemize}
	\item \textbf{Rust:} \texttt{nightly-2025-08-19}, edition 2024.
	\item \textbf{Targets:} native host execution for the prover and \texttt{riscv32imac-unknown-none-elf} for guest programs.
	\item \textbf{LatticeFold:} a fork of Nethermind's \texttt{latticefold} implementation, used for Ajtai commitments, decomposition, linearization, and NIFS folding.
	\item \textbf{Plonky3 core crates:} \texttt{p3-field}, \texttt{p3-goldilocks}, \texttt{p3-poseidon2}, \texttt{p3-merkle-tree}, \texttt{p3-commit}, \texttt{p3-matrix}, \texttt{p3-mds}, \texttt{p3-challenger}, and \texttt{p3-symmetric}.
	\item \textbf{REVM:} version \texttt{36.0.0}, used inside the \texttt{evm} guest for replaying Ethereum block execution.
\end{itemize}

\subsection{Instruction coverage}

Emulator doesn't have a full \texttt{rv32imac} instruction coverage, just
enough for running a Fibonacci program and proving a block with REVM.
And also the zkVM doesn't constraint all the supported instructions.

\paragraph{Implemented in the VM emulator:}

\begin{itemize}
	\item \textbf{RV32I control flow:} \texttt{LUI}, \texttt{AUIPC}, \texttt{JAL}, \texttt{JALR}, \texttt{BEQ}, \texttt{BNE}, \texttt{BLT}, \texttt{BGE}, \texttt{BLTU}, \texttt{BGEU}.
	\item \textbf{RV32I memory:} \texttt{LB}, \texttt{LH}, \texttt{LW}, \texttt{LBU}, \texttt{LHU}, \texttt{SB}, \texttt{SH}, \texttt{SW}.
	\item \textbf{RV32I integer arithmetic and logic:} \texttt{ADDI}, \texttt{SLTI}, \texttt{SLTIU}, \texttt{XORI}, \texttt{ORI}, \texttt{ANDI}, \texttt{SLLI}, \texttt{SRLI}, \texttt{SRAI}, \texttt{ADD}, \texttt{SUB}, \texttt{SLL}, \texttt{SLT}, \texttt{SLTU}, \texttt{XOR}, \texttt{SRL}, \texttt{SRA}, \texttt{OR}, \texttt{AND}.
	\item \textbf{RV32M:} \texttt{MUL}, \texttt{MULHU}, \texttt{DIVU}, \texttt{REMU}.
	\item \textbf{RV32A:} \texttt{LR.W}, \texttt{SC.W}, \texttt{AMOADD.W}.
	\item \textbf{Other:} \texttt{FENCE}, \texttt{ECALL}. The local compressed decoder also contains an explicit fallback for \texttt{C.SUB}, \texttt{C.XOR}, \texttt{C.OR}, and \texttt{C.AND}, which are translated to their canonical RV32I forms.
\end{itemize}

\paragraph{Unsupported:}

\begin{itemize}
	\item \textbf{RV32I:} \texttt{EBREAK}.
	\item \textbf{RV32M:} \texttt{MULH}, \texttt{MULHSU}, \texttt{DIV}, \texttt{REM}.
	\item \textbf{RV32A:} \texttt{AMOSWAP.W}, \texttt{AMOXOR.W}, \texttt{AMOAND.W}, \texttt{AMOOR.W}.
	\item \textbf{RV32A:} \texttt{AMOMIN.W}, \texttt{AMOMAX.W}, \texttt{AMOMINU.W}, \texttt{AMOMAXU.W}.
\end{itemize}

\paragraph{Constrained inside the zkVM:}

\begin{itemize}
	\item \texttt{LUI}, \texttt{AUIPC}, \texttt{JAL}, \texttt{JALR}, \texttt{BNE}, \texttt{SW}, \texttt{ADDI}, and \texttt{ADD}.
\end{itemize}

\subsection{Parameter set}

The implementation fixes the following algebraic and cryptographic parameters:

\begin{itemize}
	\item \textbf{Base field:} the Goldilocks prime field with modulus
	      $p = 2^{64} - 2^{32} + 1 = 18446744069414584321$.
	\item \textbf{Cyclotomic ring:} coefficient representation over
	      $R_q = \mathbb{F}_q[X]/(X^{24} - X^{12} + 1)$, so the ring degree is 24.
	\item \textbf{CRT/NTT form:} the ring splits into 8 NTT components over
	      $\mathbb{F}_{q^3}$, i.e. extension degree 3 and \texttt{TAU} $= 24 / 8 = 3$.
	\item \textbf{Challenge set:} short lattice challenges use coefficients in the range $[-32, 32)$.
	\item \textbf{Ajtai/LatticeFold decomposition parameters:}
	      $B = 2^{15}$, $L = 5$, $B_{\text{small}} = 2$, and $K = 15$.
	\item \textbf{Commitment module rank:} $\kappa = 32$ ring elements.
	\item \textbf{Poseidon2 commitments:} output width 4, narrow state width/rate $(8, 4)$, wide state width/rate $(16, 12)$, full rounds 8, partial rounds 22, and S-box degree 7.
	\item \textbf{Hash security estimate:} the wide Poseidon2 sponge uses capacity 4 over a 64-bit field, which gives roughly 128-bit security.
	\item \textbf{Commitment matrix width:} the Ajtai commitment uses
	      $N = \texttt{w\_size} \cdot L = 19763 \cdot 5 = 98815$ columns.
\end{itemize}

\subsection{CCS dimensions}

The resulting instance has:

\begin{itemize}
	\item witness layout size $\texttt{w\_size} = 19763$,
	\item total $z$-vector size $n = 19768$,
	\item public input size $\ell = 4$,
	\item padded row count $m = 131072 = 2^{17}$,
	\item number of matrices $t = 125$,
	\item number of multisets $q = 52$,
	\item maximal multiset degree $d = 7$,
	\item row logarithm $s = 17$,
	\item column logarithm $s' = 15$.
\end{itemize}
