\section{Implementation optimizations}

Besides the direct translation of algebraic relations into CCS constraints,
two optimizations were done in order to keep the ZkVM runnable on local
consumer-grade hardware as long as possible in development.

\subsection{\texttt{arkworks-rs} upgrade}

The LatticeFold library, and its dependencies, use the \texttt{arkworks-rs}\footnote{\url{https://github.com/arkworks-rs}}
version \texttt{v0.4.x} libraries for low level cryptography primitives. New versions
of these foundational libraries were released with optimizations and improvements.
In order to update and gain these benefits, the local fork of \texttt{latticefold}
was migrated to new version along with forking its dependencies
\texttt{cyclotomic-rings}, \texttt{stark-rings} and also migrating them to new versions.

\subsection{Reusing matrix structures}

Another optimization is that the builder groups together constraints
that share the same algebraic shape, even if they correspond to different parts
of the verifier. Instead of creating completely separate matrices and multisets
for every single row, the implementation often allocates one small family of
matrices and reuses them across many rows. For example, all product constraints
of the form "left input times right input equals intermediate product" can be
stored in the same pair of matrices and one linear
matrix for the output product. Likewise, entire families of recomposition
constraints are accumulated into shared linear matrices, differing only in the
rows that are populated. This reduces CCS size and avoids unnecessary growth in
the number of matrices and multisets.
