\section{ZkEVM}\label{sec:zkevm}

ZkEVM, as mentioned in \ref{subsec:snarkifying_execution}, is a modified EVM,
which given state and list of transactions produces next state and a proof $\pi$
attesting to correctness of that execution. The primary use case of ZkEVMs today is
enabling "ZK rollups", L2s which execute a bundle of transactions offchain,
generate a proof and submit the bundle with the proof on the L1, where only the
proof is validated. A second use case, though not used today, is to enshrine ZkEVM
into EVM as described in \ref{subsec:snarkifying_execution}.

\subsection{Challenges of EVM}

The EVM was not designed with ZK taken into consideration:

\begin{enumerate}
	\item Many EVM opcodes are complex and expensive to prove with ZK. This has
	      led to different types of EVM compatibility among ZkEVMs \cite{VitalikButerinZkEVMTypes}.
	\item 256 bit word size means that ZK systems working over prime fields must
	      include range proofs, which increases the ZkEVM complexity.
	\item It is harder to prove a stack based VM. For example, Starkware developed
	      Cairo \cite{Cairo}, a register based model in order to implement its ZkEVM,
	      which requires a custom smart contract compiler.
	\item EVM storage uses Merkle Patricia Tries with Keccak, which are not
	      ZK friendly and have huge proving overhead.
\end{enumerate}

Proving the entire EVM within a single circuit (one proof) is computationally
and economically unfeasible. Such a proof would be several megabytes in size,
making it too expensive to store on L1, let alone prove.

\subsection{Recursive proofs}

Recursive proofs combine benefits of ZKP systems with fast prover times (e.g.,
those based FRI \cite{FRI}) and systems with short proofs (like Groth16
\cite{Groth16}). The idea is to produce a proof of a knowledge of a proof.

\begin{tikzFigure}
	\begin{tikzpicture}[
			diamondNode/.style={
					diamond, draw, thick,
					minimum width=4cm,
					minimum height=0.1cm,
					align=center, font=\sffamily
				},
			proverNode/.style={
					rectangle, rounded corners=10pt, draw, thick, fill=blue!20,
					minimum width=4cm,
					minimum height=1cm,
					text width=3.5cm,
					align=center, font=\sffamily
				},
			piNode/.style={
					circle, draw, thick, fill=white,
					minimum size=1.2cm,
					align=center, font=\sffamily\Large
				},
			piPrimeNode/.style={
					circle, draw, thick, fill=green!40,
					minimum size=1.2cm,
					align=center, font=\sffamily\Large
				},
			arrow/.style={
					-Stealth, thick, draw=black!80
				},
			labelOnArrow/.style={
					font=\sffamily\small, midway, sloped
				},
			annotationText/.style={
					text=red, font=\sffamily\small, align=center
				}
		]

		\node[proverNode] (inner) {Inner prover};
		\node[proverNode, right=2.5cm of inner] (outer) {Outer prover};

		\node[diamondNode, fill=red!25, above=0.8cm of inner] (witness) {witness};
		\node[diamondNode, fill=green!25, right=1cm of witness] (public) {public inputs};

		\node[piNode, below=1.2cm of inner.south] (pi) {$\pi$};
		\node[piPrimeNode, right=2cm of outer.east] (piprime) {$\pi'$};

		\draw[arrow] (witness.south) -- (inner.north);
		\draw[arrow] (public.east) -- (outer.north);
		\draw[arrow] (public.west) -- (inner.north);

		\draw[arrow] (inner.south) to node[font=\sffamily\small, below, xshift=-1cm] {generate} (pi.north);
		\draw[arrow] (pi.east) to[out=0, in=190, looseness=1] node[font=\sffamily\small, right] {Take in as witness} (outer.west);
		\draw[arrow] (outer.east) to node[font=\sffamily\small, below] {generate} (piprime.west);

	\end{tikzpicture}
	\caption{Recursive proving}
\end{tikzFigure}

The recursive ZKP is an instantiation of 2 step Incrementally Verifiable Computation (IVC)
\cite{ValiantIVC}.

\subsection{IVC}

IVC is a cryptographic primitive for proving the correctness of an
iterated and incremental computation (such as an EVM or RISC-V microprocessor execution).

Let $F$ be a step function (e.g., representing the execution of a single
instruction). The output state of step $i$ is fed as input into step $i+1$,
along with a proof $\pi_i$ attesting to the validity of the history up to that
point.

\begin{tikzFigure}
	\begin{tikzpicture}[
			% settings for node distances
			node distance=1cm and 1.5cm, % vertical and horizontal base distance
			% Styles
			processNode/.style={ % For F1, F2, Fn
					rectangle, draw, thick, fill=blue!10,
					minimum size=1cm, % makes F nodes squarish and compact
					align=center, font=\sffamily\bfseries
				},
			dataNode/.style={ % For s0, sn, u1, u2, un
					rectangle, draw, fill=green!10,
					minimum height=0.6cm, % consistent height
					align=center,
					font=\ttfamily\small,
					inner xsep=4pt, % padding inside the node
				},
			multiLineDataNode/.style={ % specific for s0 and sn with line breaks
					dataNode,
					text width=2.2cm, % allows text to wrap
					font=\sffamily\small
				},
			instanceDataNode/.style={ % specific for instance nodes (witnesses)
					dataNode,
					minimum width=0.8cm
				},
			ellipsisNode/.style={
					text centered,
					font=\large\bfseries
				},
			arrow/.style={
					-Stealth,
					thick,
					draw=black!80
				},
			labelFont/.style={font=\sffamily\footnotesize, color=black!70}
		]

		\node[multiLineDataNode] (s0) {$z_0$\\ \textit{Initial State}};
		\node[processNode, right=0.8cm of s0] (F1) {F};
		\node[processNode, right=1.5cm of F1] (F2) {F};
		\node[ellipsisNode, right=1.2cm of F2] (dots) {...};
		\node[processNode, right=2.0cm of dots] (Fn) {F};
		\node[multiLineDataNode, right=0.8cm of Fn] (sn) {$z_n, \pi_n$\\ \textit{Final State}};

		\node[instanceDataNode, below=0.8cm of F1] (u1) {$u_1$};
		\node[instanceDataNode, below=0.8cm of F2] (u2) {$u_2$};
		\node[instanceDataNode, below=0.8cm of Fn] (un) {$u_n$};

		\draw[arrow] (s0) -- (F1);
		\draw[arrow] (u1) -- (F1);
		\draw[arrow] (F1) -- node[midway, above, labelFont] {$(z_1, \pi_1)$} (F2);
		\draw[arrow] (u2) -- (F2);
		\draw[arrow] (F2) -- node[midway, above, labelFont] {$(z_2, \pi_2)$} (dots);
		\draw[arrow] (dots) -- node[midway, above, labelFont] {$(z_{n-1}, \pi_{n-1})$} (Fn);
		\draw[arrow] (un) -- (Fn);
		\draw[arrow] (Fn) -- (sn);

	\end{tikzpicture}
	\caption{Visualization of the IVC chain \cite{ValiantIVC}}
	\label{fig:ivc_diagram}
\end{tikzFigure}

The final proof $\pi_n$ serves as a cryptographic certificate. It attests that
the prover possesses a valid sequence of private witnesses (non-deterministic
advice) $u_1, \ldots, u_n$, such that applying the transition function $F$
sequentially starting from the initial state $z_0$ results in the final public
state $z_n$. The first practical implementation of IVC was achieved using
recursive SNARKs \cite{RecursiveSNARKsIVC}.

\subsection{First generation of ZkEVMs}

In the 2022 to early 2024 first generation of ZkEVMs was successfully deployed.
Thanks to recursive proving, it was feasible for these ZkEVMs to be created.
They first prove the EVM execution, and then prove knowledge of a EVM proof inside another circuit.
This enabled parallelization of proving and reduced the size and complexity of
the final proof submitted to L1. The proving system landscape of this period was
strongly influenced by PLONK \cite{Plonk}, while recursive systems such as
Plonky2 \cite{Plonky2} and Halo2 \cite{Halo2} became practical examples of this
direction. In a broader system view, this period also laid the foundations for
the more modular zkVM architectures surveyed in \cite{SoKZkVMs}. Notable references
include Scroll, zkSync Era, Polygon ZkEVM.

\subsection{Folding schemes}

Even though recursion proofs enabled teams to build ZkEVMs, they are not without
their flaws. The prover needs to have a circuit which contains the whole verification
algorithm of another proof system. It must verify expensive evaluation proofs
for polynomial commitments.

Folding removes the need to verify a full SNARK inside the circuit, replacing
it with a cheap evaluation of a random linear combination. It
still requires a verifier circuit, but a much cheaper one.
There are no FFTs, only multi-exponentiations, which do not require large
memory overhead \cite{Nova}.

It compresses two instances into one. Folding prover will fold two instances,
with corresponding witnesses, and produce single instance, s.t., if it is correct,
it is implied that also the two original instances were correct \cite{Nova}.

Given a relation $R$, which defines valid instance-witness pairs $(u, w)$ such that $R(pp, u, w) = 1$:

\begin{itemize}
	\item $pp$ = public parameters,
	\item $u$ = instance (public inputs, commitments),
	\item $w$ = witness (private data).
\end{itemize}

For example, $R$ is an equation, $u$ are coefficients and $w$ are concrete
values that satisfy the equation.

A folding scheme for a relation $R$ is an interactive protocol where the
prover $P$ and verifier $V$ reduce two instances $u$ and $U$ into a single instance $U'$.

Let the incoming step be denoted by the pair $(u_i, w_i)$ and the running
accumulator be denoted by $(U_i, W_i)$. The folding protocol proceeds as follows:

\begin{enumerate}
	\item $P$ has $(u_i, w_i) \in R$ and $(U_i, W_i) \in R$.
	\item $V$ has $u_i$ and $U_i$ (but does not know the witnesses).
\end{enumerate}

The result of their interaction is a new folded instance $U_{i+1}$ (known to
both) and a new folded witness $W_{i+1}$ (known only to $P$), such that
$(U_{i+1}, W_{i+1}) \in R$.

Thus, instead of verifying two distinct instances ($u_i$ and $U_i$), the
verifier only needs to check the single folded instance $U_{i+1}$ at the very
end of the computation.

Folding scheme can be made non-interactive by using Fiat-Shamir heuristic \cite{FiatShamir}.

\subsection{Instantiating IVC with folding scheme}

IVC instantiation via folding schemes requires wrapping the base computation
$F$ within an augmented function $F'$, where $F$ is the "application" circuit \cite{Nova, HyperNova}.
The function $F'$ is structured to perform two primary tasks:

\begin{enumerate}
	\item \textbf{Base computation:} $F'$ executes the original state transition
	      $F(z_{i-1}, u_i) \rightarrow z_i$, where $z_{i-1}$ is the state from the
	      previous step and $u_i$ is non-deterministic advice for step $i$. This
	      corresponds to verifying the correct execution of one step of the
	      underlying computation (e.g., a single RISC-V instruction cycle).

	\item \textbf{IVC maintenance:} This task ensures the integrity of the
	      recursive proof chain. $F'$ takes as non-deterministic advice the full state
	      from the previous step: $(i, z_0, z_{i-1}, U_{i-1})$. It then performs two
	      critical checks:
	      \begin{itemize}
		      \item \textbf{Consistency check:} $F'$ re-computes the hash of the
		            input state and asserts that it matches the public input $h_{i-1}$. This
		            binds the current step's execution to the correct history.
		      \item \textbf{Recursive verification:} $F'$ executes the verifier logic
		            of a Non-Interactive Folding Scheme (NIFS). It takes the running
		            accumulator $U_{i-1}$, the instance $u_{i-1}$ (an instance
		            from the step $i-1$) and claimed new folded accumulator $U_i'$.
		            After recomputing $U_i$, the check passes if $U_i = U_i'$.
	      \end{itemize}
\end{enumerate}

Finally, $F'$ outputs a new hash $h_i$, committing to the updated state $(i+1,
	z_0, z_i, U_i)$, which serves as the public input for the next step
\cite{Nova, HyperNova}. Figure~\ref{fig:ivc-folding} illustrates this data flow.

\begin{tikzFigure}
	\begin{tikzpicture}[
			node distance=1cm and 2cm,
			accumulatorNode/.style={
					rectangle, draw, fill=orange!20,
					minimum height=0.8cm,
					align=center, font=\small
				},
			instanceNode/.style={
					rectangle, draw, fill=green!20,
					minimum height=0.8cm,
					align=center, font=\small
				},
			foldNode/.style={
					rectangle, rounded corners=5pt, draw, thick, fill=blue!20,
					minimum height=2cm,
					align=center, font=\sffamily\small
				},
			arrow/.style={ -Stealth, thick },
		]

		\node[foldNode] (Fold1) {\textbf{Step i}};
		\node[foldNode, right=of Fold1] (Fold2) {\textbf{Step i+1}};

		\node[accumulatorNode, left=of Fold1, yshift=0.5cm] (U0) {$\mathsf{U}_{i-1}$};

		\node[left=of Fold1, yshift=-0.5cm] (h0) {$h_{i-1}$};

		\node[instanceNode, below left=1cm and 0.5cm of Fold1] (u0) {$u_{i-1}$};
		\node[instanceNode, below left=1cm and 0.5cm of Fold2] (u1) {$u_{i}$};

		\node[accumulatorNode, right=of Fold2, yshift=0.5cm] (U2) {$\mathsf{U}_{i+1}$};
		\node[draw=none, right=of Fold2, yshift=-0.5cm, font=\footnotesize] (h2) {$h_{i+1}$};


		% step i inputs
		\draw[arrow] (U0.east) -- ([yshift=0.5cm]Fold1.west);
		\draw[arrow] (h0.east) -- ([yshift=-0.5cm]Fold1.west);
		\draw[arrow] (u0) -- (Fold1.225);

		% step i -> step i+1
		\draw[arrow] ([yshift=0.5cm]Fold1.east) -- node[midway, above, font=\footnotesize] {$\mathsf{U}_{i}$} ([yshift=0.5cm]Fold2.west);
		\draw[arrow] ([yshift=-0.5cm]Fold1.east) -- node[midway, above, font=\footnotesize] {$h_{i}$} ([yshift=-0.5cm]Fold2.west);

		% step i+1 inputs
		\draw[arrow] (u1) -- (Fold2.225);

		% step i+1 outputs
		\draw[arrow] ([yshift=0.5cm]Fold2.east) -- (U2.west);
		\draw[arrow] ([yshift=-0.5cm]Fold2.east) -- (h2.west);

	\end{tikzpicture}
	\caption{IVC instantiation with folding scheme \cite{Nova, HyperNova}}
	\label{fig:ivc-folding}
\end{tikzFigure}

\subsection{New generation of ZkEVMs}

Currently, the industry is moving towards ZkVMs, dominantly RISC-V ZkVMs. RISC-V is
an open-source Instruction Set Architecture (ISA) \cite{SoKZkVMs, RISCV}. RISC-V's
simpler, reduced instruction set is better suited for ZK proof computations.
Its register-based architecture is generally more efficient for proving than
the EVM's stack-based architecture. Another benefit is the mature tooling
ecosystem, debuggers, and libraries.
This allows developers to write provable programs in familiar languages like
Rust, C, and C++. RISC-V is a general purpose ISA, which opens possibilities
for proving not just blockchain applications. Due to this, a EVM written in
a language that can be compiled into RISC-V, can be executed inside a ZkVM and
together this combination creates a ZkEVM.

\subsection{Comparison}

The only sensible comparison method for ZkEVMs is to benchmark them on proving
Ethereum blocks. EthProofs \cite{EthProofs} is a portal where ZkEVM teams can
join and be measured on the most important thing, the proving times of Ethereum
blocks. As of writing this work, around 75\% of blocks are proven under 10 seconds.

The leading active systems in this comparison are STARK-based
\cite{STARK}, but their internal architecture is not uniform.
Airbender uses AIR constraint system with a DEEP-STARK/FRI pipeline.
ZisK uses PIL2 as a state machine arithmetization with a STARK prover.
Pico uses AIR with modular STARK and CircleSTARK backends. OpenVM is
also AIR-based, but organizes much of the machine through modular chips, buses,
interactions, and lookup/range check relations, and then wraps the result for
Ethereum with Halo2 on BN254. Nexus zkVM 3.0 is AIR-based and uses lookup/logup
relations over the Stwo Circle-STARK prover. SP1 is
again somewhat different. Its Hypercube stack is AIR-based, but the base proof
system is better described as a polynomial-IOP/STARK construction built
from subprotocols such as Jagged PCS and Basefold rather than as a conventional
univariate STARK pipeline. At the same time, these systems also differ in ISA
design, trace structure, memory organization, and recursion strategy, which
complicates direct performance comparison \cite{SoKZkVMs}.

Table \ref{tab:zkvm-comparison} compares selected zkVMs relevant to this
thesis. Airbender, ZisK, Pico, and OpenVM 2.0 were active on EthProofs at the
time of writing, while SP1 was listed there as inactive and Nexus zkVM was not
listed. Because these systems expose different final proof artifacts,
compressed recursive proofs in some cases and wrapped SNARKs in others, proof
size figures should be interpreted only as approximate.

\begin{table}[h]
	\centering
	\scriptsize
	\begin{tabularx}{\textwidth}{@{}l>{\centering\arraybackslash}X>{\centering\arraybackslash}X>{\centering\arraybackslash}X>{\centering\arraybackslash}X>{\centering\arraybackslash}X>{\centering\arraybackslash}X@{}}
		\toprule
		                            & \textbf{Airbender} & \textbf{ZisK} & \textbf{Pico}      & \textbf{SP1}  & \textbf{Nexus}  & \textbf{OpenVM}     \\
		\midrule
		\textbf{Organization}       & Matter Labs        & SilentSig     & Brevis             & Succinct      & Nexus           & Axiom               \\
		\midrule
		\textbf{EthProofs hardware} & 8$\times$RTX 5090  & 24 GPU        & 16$\times$RTX 5090 & n/a           & n/a             & 16$\times$ RTX 5090 \\
		\midrule
		\textbf{EthProofs time}     & 5.9s               & 8.3s          & 9.8s               & n/a           & n/a             & 4.3s                \\
		\midrule
		\textbf{Proof size}         & $\sim$1MB          & $\sim$260KB   & $\sim$1MB          & 260B / 868B   & n/a             & 277 KB              \\
		\midrule
		\textbf{ISA}                & RV32IM             & RV64IMA       & RV32IM             & RV64IM        & RV32IM          & RV32IM              \\
		\midrule
		\textbf{Constraint system}  & AIR                & PIL2          & AIR                & AIR           & AIR + logup     & AIR                 \\
		\textbf{Base zk system}     & DEEP-STARK/FRI     & STARK         & STARK              & Hypercube     & Circle STARK    & STARK + FRI         \\
		\midrule
		\textbf{Wrapper}            & SNARK              & PLONK/FFLONK  & Groth16            & Groth16 PLONK & none documented & Halo2/BN254         \\
		\bottomrule
	\end{tabularx}
	\caption{Comparison of selected zkVMs relevant to this thesis. EthProofs hardware and proving-time entries are shown where publicly available.}
	\label{tab:zkvm-comparison}
\end{table}

\pagebreak
