\chapter{Conclusion}\label{chap:conclusion}

This thesis investigated whether a general-purpose RISC-V ZkVM can be built
from lattice-based cryptographic primitives and still remain compatible with an
IVC-style proving architecture. The main result is a concrete design for such a
system. The proposed architecture combines a 32-bit RISC-V execution model,
CCS-based arithmetization, LatticeFold recursion, deferred memory checking,
and a Greyhound wrapping SNARK proof. In this sense,
the thesis provides not only a conceptual motivation for a post-quantum ZkVM,
but also a blueprint of how the individual components can be
connected into a single proving pipeline.

The implementation demonstrates that the main building blocks of this design can
be made to work together. A functional RISC-V emulator was implemented, witness
generation for the execution trace was integrated with CCS construction,
Poseidon2 was constrained inside the proving system, and the LatticeFold
IVC chain was connected to the execution layer. The experimental results show
that the current prototype is significantly slower than established zkVMs, and
profiling indicates that the dominant cost lies in multilinear-extension and
folding related algebraic computation rather than in the VM execution itself.

At the same time, the results are not purely negative. The system successfully
produced artifacts for the benchmarked execution, and a
longer \texttt{revm}-based EVM experiment continued for more than nine hours
while the IVC chain was being verified and memory usage remained bounded. This
does not yet show practical competitiveness with production zkVMs, but it does
support the central claim of the thesis: a lattice-based, non-hash-based,
post-quantum ZkVM is feasible at the architectural level, and its current
limitations appears to be mainly engineering and arithmetization issues rather
than a fundamental impossibility of the approach. In addition, although the
prototype does not yet implement the final Greyhound wrapper, the expected
proof size of such a wrapper remains attractive. Based on Greyhound's reported
evaluation proof sizes, a future final proof in the range of a few tens of
kilobytes appears plausible, which would be a strong practical property of this
design.

\section{Future Work}

Several directions for future research and development emerge from this work:

\begin{itemize}
	\item \textbf{Different computation model than CCS:} One important question is
	      whether CCS is the right arithmetization for a self-proving IVC system.
	      While CCS is expressive and convenient for some VM instructions, the
	      evaluation suggests that it may be a poor fit for recursive verifier
	      logic and for some non-linear cryptographic components. Alternative
	      models, such as AIR or lookup formulations, may be a better
	      foundation for future iterations.

	\item \textbf{SIMD and GPU acceleration for lattice operations:} A large part
	      of the current prover cost comes from lattice arithmetic, decomposition,
	      multilinear extensions, and related linear algebra. These operations are
	      regular and parallel, which makes them promising candidates for both
	      SIMD optimization and GPU acceleration.

	\item \textbf{LatticeFold+ or Neo as a backend:} Replacing the current folding
	      backend with LatticeFold+ \cite{LatticeFoldPlus} or Neo \cite{Neo} is a
	      natural next step. LatticeFold+ offers a more efficient folding design,
	      while Neo is especially interesting because it targets CCS over small
	      fields directly and introduces pay-per-bit commitments.

	\item \textbf{Algebraic and architectural optimization:} Beyond low-level
	      implementation tuning, the system may benefit from more specialized
	      lattice oriented mathematical optimizations. Another promising direction
	      is to revisit the proving architecture itself. Instead of proving steps in
	      a mostly inline sequential manner, a tree shaped proving structure may
	      allow much stronger parallelization and significantly better hardware
	      utilization.
\end{itemize}
