\section{ZkEVM}\label{sec:zkevm}

ZkEVM, as mentioned in \ref{subsec:snarkifying_execution}, is a modified EVM,
which given state and list of transactions produces next state and a proof $\pi$
attesting to correctness of that execution. Primary use case of ZkEVMs today is
enabling "ZK rollups", L2s which execute a bundle of transactions offchain,
generate a proof and submit the bundle with the proof on the L1, where only the
proof is validated. Second use case, but not used today, is to enshrine ZkEVM
into EVM as described in \ref{subsec:snarkifying_execution}.

\subsection{Challenges of EVM}

EVM was not designed with ZK taken into consideration:

\begin{enumerate}
	\item Many EVM opcodes are complex and expensive to prove with ZK. This has
		led to different types of EVM compatibility among ZkEVMs \cite{VitalikButerinZkEVMTypes}.
	\item 256 bit word size means that ZK systems working over prime fields must
		include range proofs, which increases the ZkEVM complexity.
	\item It's harder to prove stack based VM. For example, Starkware developed
		Cairo \cite{Cairo}, a register based model in order to implement its ZkEVM.
		This requires a custom smart contract compiler.
	\item EVM storage uses Merkle Patricia Tries with Keccak, which are not
		ZK friendly and have huge proving overhead.
\end{enumerate}

Proving the entire EVM within a single circuit (one proof) is computationally
and economically unfeasible. Such a proof would be several megabytes in size,
making it too expensive to store on L1, let alone prove.

\subsection{Recursive proofs}

Recursive proofs combine benefits of ZKP systems with fast prover times (e.g.,
those based FRI \cite{FRI}) and systems with short proofs (like Groth16 \cite{Groth16}). The idea is to
produce a proof of a knowledge of a proof.

\begin{tikzFigure}
	\begin{tikzpicture}[
		diamondNode/.style={
			diamond, draw, thick,
			minimum width=4cm,
			minimum height=0.1cm,
			align=center, font=\sffamily
		},
		proverNode/.style={
			rectangle, rounded corners=10pt, draw, thick, fill=blue!20,
			minimum width=4cm,
			minimum height=1cm,
			text width=3.5cm,
			align=center, font=\sffamily
		},
		piNode/.style={
			circle, draw, thick, fill=white,
			minimum size=1.2cm,
			align=center, font=\sffamily\Large
		},
		piPrimeNode/.style={
			circle, draw, thick, fill=green!40,
			minimum size=1.2cm,
			align=center, font=\sffamily\Large
		},
		arrow/.style={
			-Stealth, thick, draw=black!80
		},
		labelOnArrow/.style={
			font=\sffamily\small, midway, sloped
		},
		annotationText/.style={
			text=red, font=\sffamily\small, align=center
		}
	]

	\node[proverNode] (inner) {Inner prover};
	\node[proverNode, right=2.5cm of inner] (outer) {Outer prover};

	\node[diamondNode, fill=red!25, above=0.8cm of inner] (witness) {witness};
	\node[diamondNode, fill=green!25, right=1cm of witness] (public) {public inputs};

	\node[piNode, below=1.2cm of inner.south] (pi) {$\pi$};
	\node[piPrimeNode, right=2cm of outer.east] (piprime) {$\pi'$};

	\draw[arrow] (witness.south) -- (inner.north);
	\draw[arrow] (public.east) -- (outer.north);
	\draw[arrow] (public.west) -- (inner.north);

	\draw[arrow] (inner.south) to node[font=\sffamily\small, below, xshift=-1cm] {generate} (pi.north);
	\draw[arrow] (pi.east) to[out=0, in=190, looseness=1] node[font=\sffamily\small, right] {Take in as witness} (outer.west);
	\draw[arrow] (outer.east) to node[font=\sffamily\small, below] {generate} (piprime.west);

	\end{tikzpicture}
	\caption{Recursive proving}
\end{tikzFigure}

The recursive ZKP is an instantiation of Incrementally-Verifiable-Computation (IVC)
\cite{ValiantIVC}. IVC is a cryptographic primitive for proving the
correctness of an iterated and incremental computation (such as EVM, or RISC-V
microprocessor). The output of step $i$ of the computation ($F$) is fed as input
into step $i+1$ of the computation, along with a proof $\pi$, which proves that:

\begin{enumerate}
	\item $F(z_{i-1}, u_i) = z_i$ - proves that executing $F$ with $z_{i-1}$
		and $u_i$ correctly outputs $z_i$,
	\item $V((i-1, z_0, z_{i-1}), \pi_{i-1}) = true$ - proves that $\pi_{i-1}$ is
		a valid proof relative to the previous step of the computation.
\end{enumerate}

\begin{tikzFigure}
	\begin{tikzpicture}[
		% settings for node distances
		node distance=1cm and 1.5cm, % vertical and horizontal base distance
		% Styles
		processNode/.style={ % For F1, F2, Fn
			rectangle, draw, fill=blue!20,
			minimum size=1cm, % makes F nodes squarish and compact
			align=center
		},
		dataNode/.style={ % For s0, sn, w1, w2, wn
			rectangle, draw, fill=green!20,
			minimum height=0.5cm, % consistent height
			align=center,
			inner xsep=2pt, % padding inside the node
		},
		multiLineDataNode/.style={ % specific for s0 and sn with line breaks
			dataNode,
			text width=2.5cm % allows text to wrap
		},
		instanceDataNode/.style={ % specific for instance nodes
			dataNode,
			minimum width=1cm
		},
		ellipsisNode/.style={
			text centered,
			font=\small
		},
		arrow/.style={
			-Stealth,
			semithick
		},
		labelFont/.style={font=\footnotesize}
	]

	\node[multiLineDataNode] (s0) {$z_0$\\init state};
	\node[processNode, right=0.5cm of s0] (F1) {F};
	\node[processNode, right=of F1] (F2) {F};
	\node[ellipsisNode, right=1.5cm of F2] (dots) {...};
	\node[processNode, right=2cm of dots] (Fn) {F};
	\node[multiLineDataNode, right=0.5cm of Fn] (sn) {$(z_n, \pi_n)$\\final output};

	\node[instanceDataNode, below=of F1] (w1) {$u_1$};
	\node[instanceDataNode, below=of F2] (w2) {$u_2$};
	\node[instanceDataNode, below=of Fn] (wn) {$u_n$};

	\draw[arrow] (s0) -- (F1);
	\draw[arrow] (w1) -- (F1);
	\draw[arrow] (F1) -- node[midway, above, labelFont] {$(z_1, \pi_1)$} (F2);
	\draw[arrow] (w2) -- (F2);
	\draw[arrow] (F2) -- node[midway, above, labelFont] {$(z_2, \pi_2)$} (dots);
	\draw[arrow] (dots) -- node[midway, above, labelFont] {$(z_{n-1}, \pi_{n-1})$} (Fn);
	\draw[arrow] (wn) -- (Fn);
	\draw[arrow] (Fn) -- (sn);

	\end{tikzpicture}
	\caption{IVC \cite{ValiantIVC}}
\end{tikzFigure}

Final proof $\pi_n$ is a succinct proof that prover has $\u_1, \ldots, \u_n$,
s.t., final output $z_n$ is correct. The first implementation of IVC was done
using SNARKs \cite{RecursiveSNARKsIVC}

\subsection{First generation of ZkEVMs}

In the 2022 to early 2024 first generation of ZkEVMs was successfully deployed.
Thanks to recursive proving, it was feasible for these ZkEVMs to be created.
They first prove the EVM execution, and then prove it inside another circuit.
This enabled parallelization of proving and reduced the size and complexity of
the final proof submitted to L1. ZK systems like Plonky2 \cite{Plonky2}, or
Halo2 \cite{Halo2}, are examples of recursive ZKP systems. Notable references
include Scroll, zkSync Era, Polygon ZkEVM.

\subsection{Folding schemes}

Even though recursion proofs enabled teams to build ZkEVMs, they are not without
their flaws. Prover needs to have circuit which contains the whole verification
algorithm of another proof system. It must verify expensive evaluation proofs
for polynomial commitments.

Folding scheme takes almost all of the verification steps out of the circuit.
There are no FFTs, only multi-exponentiations, which do not require big
memory overhead. Also no embedded elliptic curve pairings are needed, because
there is no need to switch curves like in recursive proofs.

It compresses two instances into one. Folding prover will fold two instances,
with corresponding witnesses, and produce single instance, s.t., if it is correct,
it is implied that also the two original instances were correct \cite{Nova}.

Given some relations $R = \alpha(pp,\;x,\;W)$, where:
\begin{itemize}
	\item $pp$ = public parameters,
	\item $x$ = instance or public inputs,
	\item $W$ = witness.
\end{itemize}

For example: $R$ is an equation, $x$ are coefficients and $W$ are concrete
values that satisfy the equation.

A folding scheme for two relations $R$ and $R_{acc}$ (or $R_1$ and $R_2$, or
$u_i$ and $U_i$ respectively) is an interactive protocol between folding
prover $P$ and folding verifier $V$ where:

\begin{enumerate}
	\item $P$ has $(pp, x_1, w_1) \in R, (pp, x_2, w_2) \in R_{acc}$,
	\item $V$ has $(pp, x_1), (pp, x_2)$.
\end{enumerate}

And the result of their interaction is $(pp, x_3, w_3)$ (where $V$ only knows
$(pp, x_3)$ and $w_3$ is kept private for $P$). Thus instead of validating two
instances, verifier only needs to check one.

Folding scheme can be made non-interactive by using \cite{FiatShamir}.

\subsection{Instantiating IVC with folding scheme}

IVC instantiation via folding schemes requires wrapping the base computation
$F$ within an augmented function $F'$, where $F$ is the application circuit. The
function $F'$ is structured to perform two primary tasks:

\begin{enumerate}
	\item \textbf{Base Computation:} $F'$ executes the original state transition
		$F(z_{i-1}, u_i) \rightarrow z_i$, where $z_{i-1}$ is the state from the
		previous step and $u_i$ is non-deterministic advice for step $i$. This
		corresponds to verifying the correct execution of one step of the
		underlying computation (e.g., a single RISC-V instruction cycle).

	\item \textbf{IVC Maintenance:} This task ensures the integrity of the
		recursive proof chain. $F'$ takes as non-deterministic advice the full state
		from the previous step: $(i, z_0, z_{i-1}, U_{i-1})$. It then performs two
		critical checks:
		\begin{itemize}
			\item \textbf{Consistency Check:} $F'$ re-computes the hash of this
				input state and asserts that it matches the public input $h_{i-1}$. This
				binds the current step's execution to the correct history.
			\item \textbf{Recursive Verification:} $F'$ executes the verifier logic
				of a Non-Interactive Folding Scheme (NIFS). It takes the running
				accumulator $U_{i-1}$ and the instance $u_{i-1}$ (an instance
				from the step $i-1$) and outputs a new folded accumulator $U_i$.
		\end{itemize}
\end{enumerate}

Finally, $F'$ outputs a new hash $h_i$, committing to the updated state $(i+1,
z_0, z_i, U_i)$, which serves as the public input for the next step
\cite{Nova, HyperNova}. Figure~\ref{fig:ivc-folding} illustrates this data flow.

\begin{tikzFigure}
	\begin{tikzpicture}[
			node distance=1cm and 2cm,
			accumulatorNode/.style={
				rectangle, draw, fill=orange!20,
				minimum height=0.8cm,
				align=center, font=\small
			},
			instanceNode/.style={
				rectangle, draw, fill=green!20,
				minimum height=0.8cm,
				align=center, font=\small
			},
			foldNode/.style={
				rectangle, rounded corners=5pt, draw, thick, fill=blue!20,
				minimum height=2cm,
				align=center, font=\sffamily\small
			},
			arrow/.style={ -Stealth, thick },
			]

			\node[foldNode] (Fold1) {\textbf{Step i}};
			\node[foldNode, right=of Fold1] (Fold2) {\textbf{Step i+1}};

			\node[accumulatorNode, left=of Fold1, yshift=0.5cm] (U0) {$\mathsf{U}_{i-1}$};

			\node[left=of Fold1, yshift=-0.5cm] (h0) {$h_{i-1}$};

			\node[instanceNode, below left=1cm and 0.5cm of Fold1] (u0) {$u_{i-1}$};
			\node[instanceNode, below left=1cm and 0.5cm of Fold2] (u1) {$u_{i}$};

			\node[accumulatorNode, right=of Fold2, yshift=0.5cm] (U2) {$\mathsf{U}_{i+1}$};
			\node[draw=none, right=of Fold2, yshift=-0.5cm, font=\footnotesize] (h2) {$h_{i+1}$};


			% step i inputs
			\draw[arrow] (U0.east) -- ([yshift=0.5cm]Fold1.west);
			\draw[arrow] (h0.east) -- ([yshift=-0.5cm]Fold1.west);
			\draw[arrow] (u0) -- (Fold1.225);

			% step i -> step i+1
			\draw[arrow] ([yshift=0.5cm]Fold1.east) -- node[midway, above, font=\footnotesize] {$\mathsf{U}_{i}$} ([yshift=0.5cm]Fold2.west);
			\draw[arrow] ([yshift=-0.5cm]Fold1.east) -- node[midway, above, font=\footnotesize] {$h_{i}$} ([yshift=-0.5cm]Fold2.west);

			% step i+1 inputs
			\draw[arrow] (u1) -- (Fold2.225);

			% step i+1 outputs
			\draw[arrow] ([yshift=0.5cm]Fold2.east) -- (U2.west);
			\draw[arrow] ([yshift=-0.5cm]Fold2.east) -- (h2.west);

	\end{tikzpicture}
	\caption{IVC instantiation with folding scheme \cite{Nova, HyperNova}}
	\label{fig:ivc-folding}
\end{tikzFigure}

\subsection{New generation of ZkEVMs}

Nowadays, industry is moving towards ZkVMs, dominantly RISC-V ZkVMs. RISC-V is
an open sourced Instruction Set Architecture (ISA) \cite{RISCV}. RISC-V's
simpler, reduced instruction set is better suited for ZK proof computations.
Its register-based architecture is generally more efficient for proving than
the EVM's stack-based architecture. Another benefit is a mature tooling
ecosystem, including compilers like GCC and LLVM, debuggers, and libraries.
This allows developers to write provable programs in familiar languages like
Rust, C, and C++. RISC-V is a general purpose ISA, which opens possibilities
for proving not just blockchain applications. Due to this, a EVM written in
a language that can be compiled into RISC-V, can be executed inside a ZkVM and
together this combination creates a ZkEVM.

\subsubsection*{RISC-0}

First such ZkVM was RISC-0 \cite{RISC0}, a general purpose RISC-V VM. It implements
the RISC-V RV32IM specification (the RV32I base with the multiplication extension).
It uses STARKs \cite{STARK} with FRI \cite{FRI}, for the inner provers, and
the final output proof is generated with Groth16 \cite{Groth16}. Proving
process of RISC-0 is as follows:

\begin{enumerate}
	\item From execution of a program a collection of segments is collected,
	\item Each segment is proven with STARK based circuit \cite{RISC0ProofSystem},
	\item Pairs of these proofs are recursively proven until only one proof remains,
	\item This proof is proven with Groth16.
\end{enumerate}

\subsubsection*{SP1}

SP1 by Succinct is also a RISC-V RV32IM, proving programs in a recursive STARK \cite{STARK}
environment using FRI \cite{FRI}, with final proof being wrapped by Groth16
\cite{Groth16}, or PlonK \cite{Plonk} SNARK for small proof sizes \cite{SP1}.

\subsubsection*{Jolt}

Another RISC-V RV32I ZkVM. Uses a technique termed as \emph{lookup singularity}.
This technique produces circuits that perform only lookup into massive
precomputed lookup tables of size more than $2^{128}$ \cite{Jolt}. Underlying
lookup argument protocol is Lasso \cite{Lasso}.

\subsubsection*{OpenVM}

OpenVM introduces a RISC-V ZkVM with novel "no-CPU" design. This decouples
opcode implementation and let's developers build custom VM extensions. Using
FRI \cite{FRI} and DEEP-ALI \cite{DEEP-ALI}. The overall proving process follows
previous designs \cite{RISC0}, but enables aforementioned modularization \cite{OpenVM}.

\subsection{Comparison}

The only sensible comparison method for ZkEVMs is to benchmark them on proving
Ethereum blocks. However, this is not doable in this analysis because of two
reasons. First, only recently was there some effort to create standardized
benchmarks. EthProofs \cite{EthProofs} is a portal where ZkEVM teams can
join and be measured on the most important thing, the proving times of Ethereum
blocks. The average time for proving a Ethereum L1 block is (as of writing this
work) around 15.5 seconds, and median being 7.6 seconds.


