\thispagestyle{empty}
\pagenumbering{arabic}
\renewcommand*{\thepage}{C-\arabic{page}}

\chapter{Setup guide}

This guide was developed and tested on Arch Linux (kernel 6.18.24-1-lts) on an
11th Gen Intel Core i7-1165G7 @ 2.80GHz with 16 GiB RAM. For reference, the
benchmarks in this thesis were run on an Intel Xeon E312xx server with Ubuntu
24.04 and 64 GB RAM.

\section{Requirements}

The workspace is pinned to \texttt{nightly-2025-08-19}. Stable Rust is not
enough, because the workspace uses nightly-only features. The guest programs are
compiled for \texttt{riscv32imac-unknown-none-elf}, while the prover runs on the
native Linux host target (\texttt{x86\_64-unknown-linux-gnu} here).

\begin{itemize}
	\item Rust toolchain \texttt{nightly-2025-08-19}.
	\item Guest target \texttt{riscv32imac-unknown-none-elf}.
	\item About 14 GiB of free RAM for the proving run.
\end{itemize}

Downloading \texttt{rustup} is sufficient, everything else can be downloaded through it.
After installing it, run:

\begin{minted}[breaklines,obeytabs=true,tabsize=2,breaksymbolleft=]{bash}
rustup toolchain install nightly-2025-08-19
rustup target add --toolchain nightly-2025-08-19 riscv32imac-unknown-none-elf
\end{minted}

\section{How to run}

\begin{enumerate}
	\item Compile the Fibonacci guest.

	      \begin{minted}[breaklines,obeytabs=true,tabsize=2,breaksymbolleft=]{bash}
cd guests/fibonacci
cargo build --release --target riscv32imac-unknown-none-elf
\end{minted}

	      The resulting binary is stored at:

	      \begin{minted}[breaklines,obeytabs=true,tabsize=2,breaksymbolleft=]{text}
target/riscv32imac-unknown-none-elf/release/fibonacci
\end{minted}

	\item Run the zkVM prover from the repository root.

	      \begin{minted}[breaklines,obeytabs=true,tabsize=2,breaksymbolleft=]{bash}
RUST_LOG=debug cargo run --bin zkvm --features debug,latticefold/parallel --release -- --guest ./target/riscv32imac-unknown-none-elf/release/fibonacci
\end{minted}
\end{enumerate}

The proving run is memory-intensive and should be done on a machine with roughly
14 GiB of free RAM or more.
