\chapter{Introduction}

Ethereum, a decentralized blockchain platform launched in 2015, provides a
Turing-complete execution environment for smart contracts and decentralized
applications \cite{ethereumEthereumWhitepaper}. Its modular architecture
comprises a proof-of-stake consensus layer (Gasper \cite{VitalikGasper}) and
an execution layer governed by the Ethereum Virtual Machine (EVM
\cite{ethereumConsensusMechanisms}). A primary challenge for Ethereum is
scalability, as the need for all nodes to validate and re-execute every
transaction limits throughput and can centralize the network by increasing
resource demands.

"Snarkification," the integration of Zero-Knowledge Proofs (ZKPs) like SNARKs,
offers a path to mitigate these scalability issues. Initiatives such as the
'Lean consensus' \cite{LeanConsensus} aim to snarkify consensus, allowing
validators to verify compact proofs instead of entire state components.
Similarly, Zero-Knowledge Ethereum Virtual Machines (ZkEVMs) seek to make EVM
execution provable, enabling nodes to verify block execution via a single
proof. Enshrining a ZkEVM into Ethereum's Layer 1 could further enhance
scalability and simplify the creation of native rollups \cite{NativeRollups}.

Zero-Knowledge proofs allow proving a statement's truth without revealing
underlying information. These statements are about an element membership in NP
language, thus an entire computations can be proven with this primitives, for example
like computations done by the EVM, turning it into a ZkEVM. However,
developing ZkEVMs is challenging due to the EVM's original design: complex
opcodes, large word sizes requiring range proofs, its stack-based nature, and
ZK-unfriendly storage mechanisms. Early solutions utilized recursive proofs to
aggregate computation steps, leading to the first generation of Layer 2
ZkEVMs. Folding schemes further optimized this by compressing instances. The
current trend involves Zero-Knowledge Virtual Machines (ZkVMs), especially
those based on the simpler, register-based RISC-V ISA \cite{RISCV}, which is
more amenable to ZK proof generation.

While many ZKP systems rely on assumptions vulnerable to quantum attacks,
lattice-based cryptography provides a foundation for post-quantum secure ZK
solutions, with security often based on hard problems like SVP or SIS \cite{Ajtai1998}.

This thesis investigates the design and implementation of a RISC-V ZkVM using
lattice-based cryptography. A key application is proving Ethereum block
execution, contributing to the network's scalability and post-quantum
security. This work explores the practical construction of such a system and
aims to assess whether lattice-based cryptography can offer performance
comparable to classical ZKP methods while ensuring quantum resistance.

\section*{Document Structure}

This work includes the initial analytical work in Analysis
\ref{chap:analysis}. This chapter begins with an overview of Ethereum
\ref{sec:ethereum}, covering its consensus and execution mechanisms.
It then explores the concept of Snarkification of Ethereum \ref{sec:snarkification},
discussing Zero-Knowledge Proofs and their application to both the consensus
layer via the 'Lean consensus' \cite{LeanConsensus} initiative and the
execution layer through enshrined ZkEVMs. Following this, the report analyzes
ZkEVMs \ref{sec:zkevm}, examining the challenges of proving EVM execution, the
evolution of solutions from recursive proofs and folding schemes to modern
ZkVMs. Finally, the analysis introduces Lattice-based cryptography
\ref{sec:lattices}, outlining Ajtai commitments and the potential of lattices
for post-quantum zero-knowledge systems, including recent developments like
LaBRADOR, Greyhound, and LatticeFold/LatticeFold+.
